---
title: "8: Limited Dependent Variables in Panels and Survival Models"
subtitle: "Time Series and Panels"
author: "<large>J. Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "May 18, 2023"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
     

---
class: center, middle

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes the bibliography
library(RefManageR)

library(ggplot2)
library(dplyr)
library(nlme)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
#myBib <- ReadBib("assets/myBib.bib", check = FALSE)
# Note: don't forget to clear the knitr cache to account for changes in the
# bibliography.
library(DT)

usrightplotsexpanded <- read.csv("C:/Users/jnsno/OneDrive - Northwestern University/Seawright Teaching/TSCS Teaching/usrightplotsexpanded.csv")

usrightplotsexpanded2 <- read.csv("C:/Users/jnsno/OneDrive - Northwestern University/Seawright Teaching/TSCS Teaching/usrightplotsexpanded2.csv")

qog_std_ts_jan23 <- read.csv("C:/Users/jnsno/OneDrive - Northwestern University/Seawright Teaching/TSCS Teaching/qog_std_ts_jan23.csv")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

---
### Logit, Probit, Etc.


$$U^{*}_{i} = \mathbb{X}_{i} \mathbb{\beta} + \epsilon_{i}$$

$Y_{i} = 1$ if and only if:

$$U^{*}_{i} > 0$$

$$\mathbb{X}_{i} \mathbb{\beta} > - \epsilon_{i}$$


---
### Logit, Probit, Etc.

-   If $\epsilon_{i}$ has the uniform distribution, then this implies a linear probability model.

-   If $\epsilon_{i}$ has the normal distribution, then this implies a probit.

-   If $\epsilon_{i}$ has what's called the Gumbel distribution, then
    this implies a logit.



---
### Random Effects and Panel GLMs

$$U^{*}_{i,t} = \mathbb{X}_{i,t} \mathbb{\beta} + \alpha_{i} + \epsilon_{i,t}$$

$Y_{i,t} = 1$ if and only if:

$$U^{*}_{i.t} > 0$$
---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
library(lme4)

qog_std_ts_jan23$civilwar <- 1*(qog_std_ts_jan23$ht_regtype==19)
qog_std_ts_jan23$loggdp <- log(qog_std_ts_jan23$wdi_gdpcappppcon2017)

civwar.glm <- glmer(civilwar ~  p_polity2 + loggdp + al_ethnic2000  + 
    (1 | cname), data = qog_std_ts_jan23, family = binomial, control = glmerControl(optimizer = "bobyqa"),
    nAGQ = 10)
```


---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.glm)
```

---
### Fixed Effects and Panel GLMs

As $N$ goes to infinity, so do the number of estimation parameters.


---
### Conditional

Conditional MLE: estimate the substantive model conditional on the sum:
$\sum_{t=1}^{T} y_{i,t}$.

---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
library(survival)

civwar.clogit <- clogit(civilwar ~ p_polity2 + loggdp + al_ethnic2000 + strata(cname), data=qog_std_ts_jan23)
```

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.clogit)
```

---
### Unconditional

1. Get starting values for $\beta^{k}$ and $\alpha^{k}_{i}$, $i,i=1,...,N,k=0$, using a linear probability model.

2. Go into a loop and repeat the following steps until you run out of patience or the results converge.

3. Given $\beta^{k−1}$ and $\alpha^{k−1}_{i}$ compute $p^{k−1}_{i,t}$, which is the vector of probability values from the logit model for each observation. 

---
### Unconditional

4. Given $p^{k−1}_{i,t}$ compute $(\beta^{k} − \beta^{k−1})$ by running a new regression while weighting cases by their estimated $p$.

---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
library(bife)

civwar.bife <- bife(civilwar ~ p_polity2 + loggdp + al_ethnic2000 |cname , data=qog_std_ts_jan23)
```

---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.bife)
```


---
### Panel Discrete Choice or Survival Models?


---
### Event History/Survival Analysis

$f(t)$, the instantaneous failure rate


---
### Event History/Survival Analysis

$S(t) = 1 - F(t)$


---
### Event History/Survival Analysis

$h(t) = \frac{f(t)}{S(t)}$


---
### Event History/Survival Analysis

Censoring and truncation

--

Censoring: missing data where we know the range within which $S(t,i)$ falls.

Truncation: missing data where cases with values of $S(t,i)$ below a certain level disappear from the dataset.


---
### Event History/Survival Analysis

Parametric models
---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
qogsurv <- qog_std_ts_jan23 %>% 
   group_by(cname) %>% 
   mutate(
     spell = rle(civilwar)$length[first(which(!is.na(rle(civilwar)$values)))]
   )

qogsurv <- qogsurv %>% filter(year==2000)

```

---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
civwar.weibull <- survreg(Surv(spell)~p_polity2 + loggdp + al_ethnic2000, dist='weibull', data=qogsurv)
```
---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.weibull)
```

---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
civwar.exp <- survreg(Surv(spell)~p_polity2 + loggdp + al_ethnic2000, dist='exponential', data=qogsurv)
```

---
```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.exp)
```
---

### Event History/Survival Analysis

Cox proportional hazards model

--

The hazard rate, $h(t)$, is *proportional* if a one-unit increase in $X$ always increases the hazard by $e^\beta$.


---
### Event History/Survival Analysis

When the proportional-hazards condition is met, we don't need a full model for the hazard rate, and can instead use a conditional function:

$$h_{i}(t | \mathbf{x_{i}}) = h_{0}(t) exp(\beta^{T} \mathbf{x_{i}})$$
---

```{r, echo = TRUE, out.height="100%", fig.retina = 1}
civwar.coxph <- coxph(Surv(spell)~p_polity2 + loggdp + al_ethnic2000, data=qogsurv)
```

---
```{r, echo = TRUE, out.height="100%", fig.retina = 1}
summary(civwar.coxph)
```

---
```{r, echo = TRUE, out.height="100%", fig.retina = 1}
plot(survfit(civwar.coxph),xlab="Years", ylab="Proportion Not in Civil War")
```

---
```{r, echo = TRUE, out.height="100%", fig.retina = 1}
cox.zph(civwar.coxph)
```
